# Preface and Setup

:::cdi-message
- **ID:** SC-L01
- **Type:** Gateway + Setup
- **Audience:** Public
- **Theme:** What single-cell analysis is, how this guide is structured, and how to run it
:::

## What This Guide Is

Single-cell RNA-seq analysis is often described as a sequence of steps:

QC → normalization → PCA or UMAP → clustering → markers → annotation

Most confusion does not come from running those steps.
It comes from interpreting what they mean and what they do not mean.

This free guide is designed to build disciplined reasoning.

You will learn to:

- Understand the data objects (counts and metadata)
- Apply QC with explicit thresholds and tradeoffs
- Interpret embeddings and clusters carefully
- Treat marker-based labels as hypotheses, not conclusions
- Translate results into calibrated biological claims

## What This Guide Is Not

This guide is not:

- A benchmark of tools
- A promise that clustering equals cell types
- A replacement for replication and validation

It is a structured foundation that makes downstream choices easier to defend.

---

## The Reasoning Chain

We follow a reasoning chain that mirrors how interpretation should happen in practice:

Data structure → QC metrics → normalization choices → structure (PCA or UMAP) → clustering → marker evidence → calibrated claim

Each lesson adds one layer.
Later layers do not override earlier ones.

---

## Required Software

This free track is R-centric.

You need:

- R (recent version)
- Quarto
- A few R packages for plotting and basic manipulation

If you are using RStudio, both Quarto rendering and R execution work well.

---

## Project Structure

Key folders in this repository:

- `index.qmd`: cover page only
- `01-*.qmd` to `06-*.qmd`: lesson chapters
- `scripts/R/`: helper scripts (global plot theme and demo data generator)
- `data/`: small demo datasets created locally
- `docs/`: rendered site output (Quarto book)

The build is configured to output into `docs/`.

---

## Global CDI Plot Theme

This guide uses a global plotting theme so visuals stay consistent across domains.

The theme lives here:

- `scripts/R/cdi-plot-theme.R`

You will source it at the top of lessons that generate plots.

```{r source-plot-theme}
#| message: false
#| warning: false

source("scripts/R/cdi-plot-theme.R")
```

---

## Install Packages

Install packages once.

```{r install-packages}
#| eval: false

install.packages(c("ggplot2", "dplyr", "tidyr", "readr"))
```

Then load them in lessons when needed.

```{r load-packages}
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
```

---

## Generate the Demo Data

The free track uses small simulated data so the workflow is reproducible and fast.

Run the generator script once from the project root.

```{r generate-demo-data}
#| eval: false

source("scripts/R/cdi-single-cell-simulate-data.R")
```

This will create:

- `data/demo-counts.csv`
- `data/demo-metadata.csv`

If those files already exist, you can keep them.
Regenerating is fine as long as you accept that random simulation changes the numbers.

---

## Render the Book

From the project root, render with Quarto:

```bash
quarto render
```

If you want to render a single chapter during editing:

```bash
quarto render 01-preface-and-setup.qmd
```

---

## How to Use This Free Track

A practical way to move through the guide:

1. Read the concept sections
2. Run the code
3. Compare your output to the interpretation notes
4. Keep a short log of decisions (QC thresholds, normalization method, clustering resolution)

This makes your results easier to explain and easier to reproduce.

---

## Interpretation Discipline

Single-cell analysis is vulnerable to over-interpretation because:

- Cells are not independent biological replicates
- Clusters are algorithmic groupings, not ground truth
- Marker genes can be shared across states
- Batch effects can look like biology

This guide will repeatedly separate:

- What the analysis shows
- What the analysis suggests
- What the analysis cannot prove

That separation is the core skill.
