# Data Structure and QC Metrics

:::cdi-message
- **ID:** SC-L02
- **Type:** Data Foundations
- **Audience:** Public
- **Theme:** Counts matrix structure, metadata alignment, and QC reasoning
:::

## Why Structure Comes First

Single-cell analysis does not begin with UMAP.

It begins with understanding:

- What rows represent (genes)
- What columns represent (cells)
- How metadata aligns with the count matrix
- What QC metrics measure

If structure is misunderstood, every downstream result becomes fragile.

---

## Load Required Libraries

```{r load-libraries}
#| message: false
#| warning: false

source("scripts/R/cdi-plot-theme.R")

library(ggplot2)
library(dplyr)
library(readr)
```

---

## Load Demo Data

Make sure you have already generated the demo data:

```r
source("scripts/R/cdi-single-cell-simulate-data.R")
```

Now load the files.

```{r load-data}
#| message: false
#| warning: false

counts <- read.csv("data/demo-counts.csv", row.names = 1)
metadata <- read.csv("data/demo-metadata.csv", stringsAsFactors = FALSE)

dim(counts)
dim(metadata)
```

---

## Inspect the Counts Matrix

```{r inspect-counts}
counts[1:5, 1:5]
```

Interpretation:

- Rows are genes.
- Columns are cells.
- Values are raw counts.
- Counts are not normalized.
- Counts are not comparable across cells yet.

---

## Inspect Metadata

```{r inspect-metadata}
head(metadata)
str(metadata)
```

Key QC metrics:

- `nCount_RNA`: total counts per cell
- `nFeature_RNA`: number of detected genes
- `percent_mt`: proportion of mitochondrial counts

---

## Alignment Check

Counts columns must match metadata cell IDs.

```{r alignment-check}
all(colnames(counts) == metadata$cell_id)
```

If FALSE, alignment must be fixed before proceeding.

---

## Distribution of Total Counts (nCount_RNA)

```{r plot-ncount}
ggplot(metadata, aes(x = nCount_RNA)) +
  cdi_geom_histogram(bins = 40, colored = TRUE) +
  cdi_scale_histogram_fill() +
  labs(
    title = "Distribution of total counts per cell",
    subtitle = "Cells with extremely low counts may be poor quality",
    x = "nCount_RNA",
    y = "Number of cells"
  ) +
  cdi_theme()
```

Improved Interpretation:

Most cells fall within a relatively narrow range of total counts.

There is no cluster of extremely low-count cells, which reduces the likelihood of empty droplets.

A modest right tail exists. This could represent:
- Cells with genuinely higher RNA content
- Potential doublets (two cells captured together)

At this stage, there is no justification to remove cells purely based on nCount_RNA.

QC thresholds should not be applied simply because they are common in tutorials.

---

## Distribution of Detected Genes (nFeature_RNA)

```{r plot-nfeature}
ggplot(metadata, aes(x = nFeature_RNA)) +
  cdi_geom_histogram(bins = 40, colored = TRUE) +
  cdi_scale_histogram_fill() +
  labs(
    title = "Distribution of detected genes per cell",
    subtitle = "Low-feature cells are often low quality",
    x = "nFeature_RNA",
    y = "Number of cells"
  ) +
  cdi_theme()
```

Improved Interpretation:

The distribution is narrow and largely unimodal.

There is no obvious subpopulation of extremely low-feature cells.

This suggests that technical dropout is not dominating the dataset.

Higher-feature cells could reflect:
- Larger or more transcriptionally active cells
- Doublets
- Or specific biological states

Again, no immediate filtering decision is justified.

---

## Mitochondrial Percentage (percent_mt)

```{r plot-percent-mt}
ggplot(metadata, aes(x = percent_mt)) +
  cdi_geom_histogram(bins = 40, colored = TRUE) +
  cdi_scale_histogram_fill() +
  labs(
    title = "Mitochondrial percentage per cell",
    subtitle = "High mitochondrial proportion may indicate stressed or dying cells",
    x = "percent_mt",
    y = "Number of cells"
  ) +
  cdi_theme()
```

Improved Interpretation:

The mitochondrial percentage shows clear bimodality.

One group of cells has low mitochondrial content (~2–4%), while another shows substantially higher values (~14–18%).

This pattern suggests:
- Either a stressed cell subpopulation
- Or a batch-driven difference

Before filtering, we must check:

- Is high percent_mt enriched in one batch?
- Does removing these cells remove a biological cluster?
- Are these cells coherent in downstream PCA space?

Blindly applying a "5% mitochondrial cutoff" would eliminate an entire subpopulation.

---

## QC Decisions Should Be Tested

QC filtering is a hypothesis.

Before applying thresholds, test whether:

- High percent_mt cells cluster together
- They are enriched in a specific batch
- They align with a specific cell type

If filtering removes a coherent biological signal, it may be inappropriate.

QC is not about cleaning data until it looks nice.

It is about controlling technical noise while preserving biological structure.

---

## What This Lesson Established

You now understand:

- The structure of a single-cell count matrix
- The role of metadata
- Core QC metrics
- Why QC thresholds require reasoning, not memorization

Next, we move to normalization and feature selection.
