# Dimensionality Reduction and Clustering

:::cdi-message
- **ID:** SC-L04
- **Type:** Structure Discovery
- **Audience:** Public
- **Theme:** How selected variance becomes structure in reduced space
:::

---

## Why This Lesson Matters

In Lesson 3, we learned:

- PCA amplifies variance.
- Feature selection determines which variance is amplified.
- Highly variable genes shape downstream embeddings.

Now we allow that selected variance to define structure.

Dimensionality reduction does not create biology.  
It compresses variance into interpretable axes.

---

# Load Data and Recompute Variable Genes

```{r load-data}
#| message: false
#| warning: false

source("scripts/R/cdi-plot-theme.R")

library(ggplot2)
library(dplyr)

counts_df <- read.csv("data/demo-counts.csv", check.names = FALSE)
rownames(counts_df) <- counts_df[, 1]
counts_df <- counts_df[, -1, drop = FALSE]
counts <- as.matrix(counts_df)
storage.mode(counts) <- "numeric"

metadata <- read.csv("data/demo-metadata.csv", stringsAsFactors = FALSE)

library_size <- colSums(counts)
norm_counts <- sweep(counts, 2, library_size, "/") * 10000
log_counts <- log1p(norm_counts)

gene_vars <- apply(log_counts, 1, var)
var_order <- order(gene_vars, decreasing = TRUE)
variable_genes <- rownames(log_counts)[var_order[1:200]]
```

---

# Principal Component Analysis

```{r pca}
#| message: false
#| warning: false

pca_input <- t(log_counts[variable_genes, ])

pca_res <- prcomp(pca_input, center = TRUE, scale. = FALSE)

pca_df <- data.frame(
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2],
  percent_mt = metadata$percent_mt,
  batch = metadata$batch
)

summary(pca_res)$importance[2, 1:5]
```

---

# PCA Colored by Mitochondrial Percentage

```{r pca-mito-plot}
ggplot(pca_df, aes(x = PC1, y = PC2, color = percent_mt)) +
  geom_point(size = 2.2, alpha = 0.85) +
  scale_color_viridis_c(option = "viridis") +
  labs(
    title = "PCA colored by mitochondrial percentage",
    subtitle = "Technical structure may align with selected variance",
    x = "PC1",
    y = "PC2",
    color = "percent_mt"
  ) +
  cdi_theme()
```

## Interpretation

If percent_mt aligns with PC1 or PC2:

- Mitochondrial-associated genes influenced the selected variance.
- Technical variation contributes to visible structure.

Structure is diagnostic, not automatically biological.

---

# PCA Colored by Batch

```{r pca-batch-plot}
ggplot(pca_df, aes(x = PC1, y = PC2, color = batch)) +
  geom_point(size = 2.2, alpha = 0.85) +
  labs(
    title = "PCA colored by batch",
    subtitle = "Batch-driven separation reflects structured variance",
    x = "PC1",
    y = "PC2",
    color = "Batch"
  ) +
  cdi_theme()
```

## Interpretation

If batch drives separation:

- Batch influences variable genes, or  
- True biology differs between batches.

Structure alone cannot resolve this.

---

# Clustering in Reduced Space

```{r clustering}
set.seed(123)
clusters <- kmeans(pca_res$x[, 1:2], centers = 3)$cluster
pca_df$cluster <- factor(clusters)
```

---

# PCA with Cluster Assignments

```{r pca-cluster-plot}
ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2.2, alpha = 0.9) +
  labs(
    title = "PCA with cluster assignments",
    subtitle = "Clusters reflect structure in selected variance space",
    x = "PC1",
    y = "PC2",
    color = "Cluster"
  ) +
  cdi_theme()
```

## What Clusters Represent

Clusters represent cells similar in selected variance space.

They are the compression of selected variance, not biological labels.

---

# Mitochondrial Percentage by Cluster

```{r mito-by-cluster}
ggplot(pca_df, aes(x = cluster, y = percent_mt, fill = cluster)) +
  geom_boxplot(alpha = 0.8) +
  labs(
    title = "Mitochondrial percentage by cluster",
    subtitle = "Cluster-level QC alignment must be evaluated",
    x = "Cluster",
    y = "percent_mt"
  ) +
  cdi_theme()
```

## Interpretation

If one cluster has elevated mitochondrial percentage:

- Structure is partially technical.
- Marker interpretation must be calibrated.
- Biological claims must be constrained.

Structure is variance geometry.  
Meaning requires calibration.

---

# What This Lesson Established

You now understand:

- PCA amplifies selected variance.
- Technical metrics can align with structure.
- Clusters reflect compressed variance space.
- QC must be evaluated at cluster level.

Next lesson: from clusters to marker genes.
