# Lesson 4: Dimensionality Reduction and Clustering

:::cdi-message
- **ID:** SC-L04
- **Type:** Structure Modeling
- **Audience:** Public
- **Theme:** PCA, technical structure checks, clustering logic, and interpretation discipline
:::

## From Variable Genes to Structure

In the previous lesson, we:

- Normalized counts
- Log-transformed expression
- Selected highly variable genes

Now we use those genes to explore structure.

Dimensionality reduction does not create biology.
It reveals structure already present in the data.

---

## Load Data and Prepare Matrix

```{r load-data}
#| message: false
#| warning: false

source("scripts/R/cdi-plot-theme.R")

library(ggplot2)
library(dplyr)
library(readr)

counts <- read.csv("data/demo-counts.csv", row.names = 1)
metadata <- read.csv("data/demo-metadata.csv", stringsAsFactors = FALSE)

library_size <- colSums(counts)
norm_counts <- sweep(counts, 2, library_size, "/") * 10000
log_counts <- log1p(norm_counts)

gene_vars  <- apply(log_counts, 1, var)
var_order <- order(gene_vars, decreasing = TRUE)
variable_genes <- rownames(log_counts)[var_order[1:200]]

mat <- t(log_counts[variable_genes, ])
```

---

## Principal Component Analysis (PCA)

```{r run-pca}
#| message: false
#| warning: false

pca <- prcomp(mat, center = TRUE, scale. = TRUE)

pc_df <- data.frame(
  cell_id = metadata$cell_id,
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  percent_mt = metadata$percent_mt,
  batch = metadata$batch
)
```

---

## PCA Colored by Mitochondrial Percentage

```{r plot-pca-mt}
#| message: false
#| warning: false

ggplot(pc_df, aes(x = PC1, y = PC2, color = percent_mt)) +
  geom_point(size = 2, alpha = 0.85) +
  scale_color_viridis_c() +
  labs(
    title = "PCA colored by mitochondrial percentage",
    subtitle = "Technical structure may align with high percent_mt",
    x = "PC1",
    y = "PC2"
  ) +
  cdi_theme()
```

### Interpretation

**Technical structure is driving separation.**

PC1 is strongly aligned with mitochondrial percentage.
This indicates that the dominant axis of variation is not purely biological.

Cells with high percent_mt are separating in reduced space, suggesting:

- Technical stress
- Apoptotic or dying cells
- Capture inefficiency

If clustering were performed without QC awareness, this axis could be mistaken for a biological subtype.

In single-cell analysis, the first question after PCA is not:

"What cell type is this?"

It is:

"Is this biology, or is this technical structure?"

---

## PCA Colored by Batch

```{r plot-pca-batch}
#| message: false
#| warning: false

ggplot(pc_df, aes(x = PC1, y = PC2, color = batch)) +
  geom_point(size = 2, alpha = 0.85) +
  labs(
    title = "PCA colored by batch",
    subtitle = "Batch-driven separation indicates technical variation",
    x = "PC1",
    y = "PC2"
  ) +
  cdi_theme()
```

### Interpretation

**Batch does not dominate structure.**

Batches are intermingled within clusters.
There is no strong batch-specific axis.

This suggests:

- Minimal batch effect
- Reduced risk of confounding

Here, mitochondrial QC signal is stronger than batch signal.

That distinction matters.

---

## Clustering (K-means Example)

Clustering assigns cells to algorithmic groups.
Clusters are not automatically cell types.

```{r kmeans-clustering}
#| message: false
#| warning: false

set.seed(123)
kmeans_res <- kmeans(pc_df[, c("PC1", "PC2")], centers = 3)

pc_df$cluster <- as.factor(kmeans_res$cluster)
```

---

## PCA with Cluster Labels

```{r plot-pca-cluster}
#| message: false
#| warning: false

ggplot(pc_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2, alpha = 0.85) +
  labs(
    title = "PCA with cluster assignments",
    subtitle = "Clusters reflect structure in reduced space",
    x = "PC1",
    y = "PC2"
  ) +
  cdi_theme()
```

### Interpretation

**One cluster is likely QC-driven.**

The clustering algorithm did not discover biology.
It discovered structure.

One cluster corresponds closely to the high percent_mt region, suggesting:

- Stressed cells
- Low-quality cells
- A technical axis captured by PCA

The other clusters are more biologically plausible populations.

Dimensionality reduction and clustering do not create meaning.
They reveal structure.
Interpretation assigns meaning.

---

## QC Check: percent_mt by Cluster

This plot tests whether mitochondrial percentage differs systematically across clusters.

```{r boxplot-percent-mt-by-cluster}
#| message: false
#| warning: false

ggplot(pc_df, aes(x = cluster, y = percent_mt)) +
  geom_boxplot(alpha = 0.9, outlier.alpha = 0.6) +
  labs(
    title = "Mitochondrial percentage by cluster",
    subtitle = "Large shifts indicate clusters may be QC-driven",
    x = "Cluster",
    y = "percent_mt"
  ) +
  cdi_theme()
```

### Interpretation

If one cluster has substantially higher percent_mt, it is a warning signal.

It suggests the cluster may reflect:

- Cell stress
- Low-quality capture
- Technical separation

This does not force filtering.
It forces calibration.

---

## Structural Calibration Before Biological Annotation

Before labeling clusters as biological identities, rule out:

- QC-driven clusters
- Mitochondrial bias
- Library size effects
- Batch artifacts

In this dataset:

- One cluster is likely technical.
- Two clusters are biologically plausible.

That calibration step prevents false biological narratives.

---

## What This Lesson Established

You now understand:

- How PCA reveals structure
- How technical variables can align with structure
- Why clusters are hypotheses, not truths
- How to test whether a cluster is driven by QC metrics

Next, we move to marker genes and annotation discipline.
