# Lesson 5: Marker Genes and Annotation Discipline

:::cdi-message
- **ID:** SC-L05
- **Type:** Differential Signal
- **Audience:** Public
- **Theme:** Identifying marker genes and separating clustering from biological identity
:::

## From Clusters to Markers

Clustering groups cells based on expression similarity.

Marker analysis asks:

Which genes distinguish one cluster from the others?

Clusters are algorithmic.
Markers provide biological evidence.

Marker detection is still a statistical model, not proof of identity.

---

## Load Data and Recreate Objects

```{r load-and-prepare}
#| message: false
#| warning: false

source("scripts/R/cdi-plot-theme.R")

library(ggplot2)
library(dplyr)
library(readr)

# Read counts as numeric matrix (avoid character columns)
counts_df <- read.csv("data/demo-counts.csv", check.names = FALSE)
rownames(counts_df) <- counts_df[, 1]
counts_df <- counts_df[, -1, drop = FALSE]
counts <- as.matrix(counts_df)
storage.mode(counts) <- "numeric"

metadata <- read.csv("data/demo-metadata.csv", stringsAsFactors = FALSE)

# Normalize + log transform
library_size <- colSums(counts)
norm_counts <- sweep(counts, 2, library_size, "/") * 10000
log_counts <- log1p(norm_counts)

# Variable genes
gene_vars <- apply(log_counts, 1, var)
var_order <- order(gene_vars, decreasing = TRUE)
variable_genes <- rownames(log_counts)[var_order[1:200]]

# PCA for clustering
mat <- t(log_counts[variable_genes, ])
pca <- prcomp(mat, center = TRUE, scale. = TRUE)

pc_df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2]
)

set.seed(123)
kmeans_res <- kmeans(pc_df[, c("PC1", "PC2")], centers = 3)
clusters <- as.factor(kmeans_res$cluster)
table(clusters)
```

---

## Simple Marker Detection (Cluster 1 vs Others)

For teaching, we use a per-gene two-sample t-test on log-normalized expression.

In real analyses, you should use models designed for single-cell data and account for confounding, library size, and zero inflation.

```{r marker-test}
#| message: false
#| warning: false

cluster_id <- "1"

group1 <- which(clusters == cluster_id)
group2 <- which(clusters != cluster_id)

marker_results <- data.frame(
  gene = rownames(log_counts),
  logFC = NA_real_,
  pvalue = NA_real_
)

for (i in seq_len(nrow(log_counts))) {
  g1 <- as.numeric(log_counts[i, group1])
  g2 <- as.numeric(log_counts[i, group2])

  # Effect size (mean difference)
  marker_results$logFC[i] <- mean(g1, na.rm = TRUE) - mean(g2, na.rm = TRUE)

  # Some genes can be nearly constant; t.test may warn. We treat failures as NA.
  tt <- tryCatch(
    suppressWarnings(t.test(g1, g2)),
    error = function(e) NULL
  )

  marker_results$pvalue[i] <- if (is.null(tt)) NA_real_ else tt$p.value
}

marker_results$padj <- p.adjust(marker_results$pvalue, method = "BH")

# Define "significant" markers for visualization
marker_results$significant <- !is.na(marker_results$padj) & marker_results$padj < 0.05
```

---

## Volcano Plot

```{r volcano-plot}
#| message: false
#| warning: false

volcano_df <- marker_results %>%
  mutate(neglog10_padj = -log10(padj))

sig_cutoff <- -log10(0.05)

ggplot(volcano_df, aes(x = logFC, y = neglog10_padj)) +
  geom_point(aes(color = significant), alpha = 0.75) +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "#036281")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = sig_cutoff, linetype = "dashed") +
  labs(
    title = "Marker gene volcano plot (Cluster 1)",
    subtitle = "Significant genes highlighted after multiple testing correction",
    x = "Mean difference in log expression (Cluster 1 minus others)",
    y = "-log10(adjusted p-value)"
  ) +
  cdi_theme()
```

### Interpretation

This volcano plot shows three regimes:

1. **Central cloud near logFC = 0**  
   Genes with minimal difference between clusters. These form the null background.

2. **Right tail (positive logFC, high significance)**  
   Genes strongly higher in Cluster 1. Large effect sizes align with strong evidence.

3. **Left tail (negative logFC)**  
   Genes depleted in Cluster 1 relative to others.

What is striking here is not only statistical significance.

It is structural coherence.

The most significant genes are also those with large effect sizes.
That alignment suggests Cluster 1 is genuinely distinct in expression space.

However, this does not yet prove biological identity.
It confirms expression differentiation.

---

## Statistical Strength Does Not Equal Biological Identity

This plot confirms:

- Cluster 1 is transcriptionally distinct
- The separation is not subtle
- The signal survives multiple testing correction

But this does not tell us:

- Whether this cluster represents a known cell type
- Whether it reflects activation state
- Whether it reflects stress
- Whether it is a QC artifact

Remember from Lesson 4:

One cluster aligned with high percent_mt.

If this volcano plot corresponds to that cluster, we may be detecting stress-response genes.

Strong signal does not guarantee meaningful biology.

It guarantees difference.

---

## Top Marker Table

```{r top-markers}
#| message: false
#| warning: false

marker_results %>%
  arrange(padj) %>%
  select(gene, logFC, pvalue, padj) %>%
  head(10)
```

Interpretation:

These genes show the strongest statistical separation under this simple test.

Biological annotation requires:

- Known gene function and specificity
- Consistency across datasets or references
- Plausibility given the experimental context

---

## Critical Distinction

Clustering answers:

Which cells group together?

Marker analysis answers:

Which genes differentiate those groups?

Annotation answers:

What biological identity best explains those genes?

These are three separate steps.

---

## Annotation Is a Hypothesis

Even strong markers:

- May reflect activation state rather than cell type
- May be influenced by QC structure
- May reflect stress responses

Annotation should be framed as:

"Cluster 1 is consistent with X-like cells based on its marker profile."

Not:

"Cluster 1 is definitively X."

---

## What This Lesson Established

You now understand:

- How to compute simple marker statistics without type issues
- How to interpret volcano plots using both effect size and evidence
- Why strong signal does not automatically imply cell identity
- Why annotation is inference, not proof

Next, we translate clusters and markers into calibrated biological claims.
